Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is a merged and final version of two conference papers, \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}, with a unified and streamlined presentation.}

### 2020-12-20 journal paper/content/sec-intro.tex
Minor edits.

In this paper, we study the interplay between \exploration and \competition.%
\footnote{\Ie we add \competition to the standard exploration-exploitation tradeoff studied in multi-armed bandits.}
Platforms that engage in exploration typically need to compete against one another. Most importantly, platforms compete for users, who benefit them in two ways:
generating revenue and providing data for learning. This creates a tension:
%between \exploration and \competition.
while exploration may be essential for improving the service tomorrow, it may degrade quality \emph{today}, in which case some of the users can leave and there will be fewer users to learn from. This may create a ``data feedback loop" when the platform's performance further degrades relative to competitors who keep learning and improving from \emph{their} users, and so forth. Taken to the extreme, such dynamics may cause a ``death spiral" effect when the vast majority of customers eventually switch to competitors.

@@

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a multi-armed bandit algorithm; as such, it can dynamically adjust itself to the observed rewards, but not to the \asedit{competitor's choices, rewards, or market share}.
%\footnote{\gaedit{In particular, the algorithm does not respond to neither the competitors' rewards (which it does not observe) nor its current market share.}} 
This modeling choice reflects the reality of industrial applications, which build on a huge body of knowledge in machine learning; more on this in Section~\ref{sec:discussion}.
%Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical.

>>>>

In this paper, we study the interplay between \exploration and \competition.%
\footnote{\Ie we add \competition to the standard exploration-exploitation tradeoff studied in multi-armed bandits.}
Platforms that engage in exploration typically need to compete against one another. Most importantly, platforms compete for users, who benefit them in two ways:
generating revenue and providing data for learning. This creates a tension:
%between \exploration and \competition.
while exploration may be essential for improving the service tomorrow, it may degrade the service quality \emph{today}, in which case some of the users can leave and there will be fewer users to learn from. This may create a ``data feedback loop" when the platform's performance further degrades relative to competitors who keep learning and improving from \emph{their} users, and so forth. Taken to the extreme, such dynamics may cause a ``death spiral" effect when the vast majority of customers eventually switch to competitors.

@@

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a multi-armed bandit algorithm. As such, it can dynamically adjust itself to the observed rewards, but not to \asedit{any aspect of competition (incl. the market share and the competitor's choices or rewards).}
%\footnote{\gaedit{In particular, the algorithm does not respond to neither the competitors' rewards (which it does not observe) nor its current market share.}} 
This modeling choice reflects the reality of industrial applications, which build on a huge body of knowledge in machine learning; more on this in Section~\ref{sec:discussion}.
%Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical.

### 2020-12-20 journal paper/journal_main.tex
Minor edits.

\vspace{-5mm}
\begin{abstract}
Most online platforms strive to learn from interactions with users, and many engage in \emph{exploration}: making potentially suboptimal choices for the sake of acquiring new information. We initiate a study of the interplay between \emph{exploration} and \emph{competition}: how such platforms balance the exploration for learning and the competition for users. Here users play three distinct roles: they are customers that generate revenue, they are sources of data for learning, and they are self-interested agents which choose among the competing platforms.

We consider a stylized duopoly model in which two firms face the same problem instance of multi-armed bandits. Users arrive one by one and choose between the two firms, so that each firm makes progress on its bandit instance only if it is chosen. Through a mix of theoretical results and numerical simulations, we study whether and to what extent competition incentivizes the adoption of better bandit algorithms, and whether it leads to welfare increases for users. We find that stark competition induces firms to commit to a ``greedy" bandit algorithm that leads to low welfare. However, weakening competition by providing firms with some ``free" users incentivizes better exploration strategies and increases welfare.  We investigate two channels for weakening the competition: relaxing the rationality of users and giving one firm a first-mover advantage. Our findings are closely related to the ``competition vs. innovation" relationship, and elucidate the first-mover advantage in the digital economy.

>>>>

\vspace{-5mm}
\begin{abstract}
Most online platforms strive to learn from interactions with users, and many engage in \emph{exploration}: making potentially suboptimal choices for the sake of acquiring new information. We \asedit{study} the interplay between \emph{exploration} and \emph{competition}: how such platforms balance the exploration for learning and the competition for users. Here users play three distinct roles: they are customers that generate revenue, they are sources of data for learning, and they are self-interested agents which choose among the competing platforms.

We consider a stylized duopoly model in which two firms face \asedit{the same multi-armed bandit problem}. Users arrive one by one and choose between the two firms, so that each firm makes progress on its bandit instance only if it is chosen. Through a mix of theoretical results and numerical simulations, we study whether and to what extent competition incentivizes the adoption of better bandit algorithms, and whether it leads to welfare increases for users. We find that stark competition induces firms to commit to a ``greedy" bandit algorithm that leads to low welfare. However, weakening competition by providing firms with some ``free" users incentivizes better exploration strategies and increases welfare.  We investigate two channels for weakening the competition: relaxing the rationality of users and giving one firm a first-mover advantage. Our findings are closely related to the ``competition vs. innovation" relationship, and elucidate the first-mover advantage in the digital economy.
